Constraints and Opportunities for Detecting Land Surface Phenology in Drylands

### 2021-06-30 manuscript/main.tex
add line numbers
changes to text
Addresses most reviewer comments.

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{setspace}
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{subcaption}
\usepackage{amsmath}

@@

%%%%%% Abstract %%%%%%
\begin{abstract}
Land surface phenology, the tracking of seasonal productivity via satellite remote sensing, enables global scale tracking of ecosystem processes, but its utility is limited in some areas. In dryland ecosystems low vegetation cover can cause the growing season vegetation index (VI) to be indistinguishable from the dormant season VI, making phenology extraction impossible. Here, using simulated data and multi-temporal UAV imagery of a desert shrubland, we explore the feasibility of detecting LSP with respect to fractional vegetation cover, plant functional types, and VI uncertainty. We found that plants with distinct VI signals, such as deciduous shrubs with a high leaf area index, require at least 30-40\% fractional cover on the landscape to consistently detect pixel level phenology with satellite remote sensing. Evergreen plants, which have lower VI amplitude between dormant and growing seasons, require considerably higher cover and can have undetectable phenology even with 100\% vegetation cover. We also found that even with adequate cover, biases in phenological metrics can still exceed 20 days, and can never be 100\% accurate due to VI uncertainty from shadows, sensor view angle, and atmospheric interference. Many dryland areas do not have detectable LSP with the current suite of satellite based sensors. Our results showed the feasibility of dryland LSP studies using high-resolution UAV imagery, and highlighted important scale effects due to within canopy VI variation. Future sensors with sub-meter resolution will allow for identification of individual plants and are the best path forward for studying large scale phenological trends in drylands.

@@

Landscape patterns in seasonal productivity [hereafter ‘land surface phenology’ (LSP)] enable ecosystem scale tracking of the drivers and consequences of a changing climate. Satellite sensor-derived vegetation indices (VIs) track the progression of green vegetation throughout the year, and from this time series the seasonal transitions between the dormant and growing seasons are derived \cite{zhang-liu2018, friedl2006, jonsson2004, sakamoto2010, gao2017, gray2019, bolton2020, zeng-wardlow2020}. Numerous studies measure the short- and long-term LSP trends, linking them to drivers such as weather and climate, land cover change, and disturbances \cite{piao-wang2020}. Yet the use of LSP is limited outside regions with adequate vegetation cover and plants with distinct seasonal change. Temperate deciduous forests have the most distinct LSP signal, while in other vegetation types LSP is difficult to discern due to combinations of low signal-to-noise ratio between growing and dormant season VI, high snow cover, and low vegetation cover \cite{richardson2018b, white2009}.

In dryland ecosystems low vegetation cover is the primary limitation in detecting land surface phenology \cite{ma2013}. High soil fractional cover can cause the growing season VI to be indistinguishable from the dormant season VI, making phenology extraction impossible. Additionally, some areas with adequate vegetation cover are dominated by evergreen vegetation, thus can have a growing season VI too low to reliably detect seasonal transitions \cite{walker2015}. Even in areas with high vegetation cover and plants with distinct signals, phenology may be impossible to detect some years due little to no plant productivity from inadequate precipitation. These limitations cause drylands to be excluded in many large scale LSP analyses \cite{jeong2011, zhu2012, shen-zhang2014, xin-broich2015, garonna2016} Most studies focusing on drylands evaluate aggregate or peak annual VI as opposed to distinct seasonal transitions \cite{walker2015, hou-du2019}, and even then can occasionally have inconclusive results \cite{fensholt2012, andela2013}.

Here we explore key attributes of VI time series and how they affect reliable detection of LSP in dryland ecosystems. We use a spectral mixing model to simulate LSP with varying ranges of plant fractional cover, seasonal VI amplitude, and VI uncertainty to determine the feasibility of detecting LSP in different scenarios. We then use high spatial resolution, multi-temporal unmanned aerial vehicle (UAV) imagery to verify some of these relationships. Finally we discuss the feasibility of detecting dryland LSP currently and in the future.

@@

Here we explore how variation in the three components in Eq. 1 (i.e., ranges of plant fractional cover, seasonal VI amplitude, and VI uncertainty) affect the ability to detect a distinct phenological signal in sparsely vegetated areas. We start with an idealized annual VI curve derived from a double sigmoid model to represent the VI of single plant canopy (Figure 1, \cite{elmore2012}). The curve’s amplitude (i.e., the difference between dormant and peak VI values) reflects a combination of plant functional type and leaf area index (LAI). The seasonal amplitude of non-evergreen plants is driven by LAI up to an LAI of approximately 2, while evergreen plants have little to no seasonal amplitude \cite{gamon1995, carlson1997}. Higher LAI will generally lead to higher amplitudes, with this effect being more prominent in deciduous plants than evergreen plants \cite{baghzouz2010, smith2019}. The leaves of senesced grasses have distinctly lower NDVI than green leaves, thus are expected to behave similarly to deciduous plants \cite{huete1987}.

@@

For the simulation analysis we repeated this process while varying the three different factors from Eq. 1: 1) the fractional vegetation cover ($Cover_{veg}$), using values from 0-100\%, 2) the amplitude of $VI_{veg}$ from dormant to peak, adjusted using a parameter in the double sigmoid model, and 3) the size of the error term $\varepsilon$ using values of 0.01 and 0.04 for the standard deviation of the gaussian noise. For each combination we repeated the process 100 times. A low growing season VI relative to the dormant season is one of the main challenges in dryland LSP \cite{smith2019}. Phenological extraction methods account for this in several ways. A common approach is discarding pixels which do not exceed a threshold of 0.1 VI units between the dormant season and peak growth (i.e., a minimum $VI_{pixel}$ amplitude of 0.1; \cite{white2009, jeong2011, garonna2016, gray2019}), while others have used statistical power methods of the time-series to filter out pixels with low signal-to-noise \cite{dannenberg2017}. Here we evaluate the detectability of LSP in two ways: 1) The frequency at which the $VI_{pixel}$ meets or exceeds an amplitude of 0.1 VI units, and 2) mean absolute error (MAE) in the estimated phenological metrics start of season (SOS), end of season (SOS), and peak.

\subsection{UAV Study Area}

@@

\subsection{UAV Acquisition and Processing}

@@

Using this time series of NDVI rasters we developed a sampling routine to test the detectability of phenology with respect to fractional vegetation cover similar to the simulation study. We randomly placed 12,000 simulated pixels across the site with sizes of 2, 4, 6, 8, or 16 meters. We then extracted the average NDVI within each one for each image date. This random placement allowed us to produce 12,000 annual NDVI time series representing a range of fractional cover for honey mesquite across several pixel resolutions. The average shrub diameter is approximately 8m, so multiple pixel sizes allow us to test both L- and H-resolution LSP detection \cite{strahler1986}. As in the simulation study, each annual time series was fit with a cubic smoothing spline and we then calculated whether the smoothed time series exceeded a 0.1 NDVI threshold and extracted phenology metrics using a 10\% of relative max threshold method and the curvature method as described below.

@@

\subsection{Detection Methods and Analysis}

Many remote sensing algorithms have been developed in recent decades for estimating transition dates from VI timeseries. They can be categorized in the threshold, curvature, trend, and priori curve-based approach \cite{gao-zhang2021}. The threshold-based approach uses the predefined threshold of amplitude of VI (difference between base and maximum VI) to detect SOS, EOS, and other phenological metrics \cite{jonsson2004, gray2019, bolton2020}. The curvature-based approach detects the inflection points that show the quickest changes in the VI time series \cite{zhang-friedl2003, zhang-liu2018, gao2017}. The trend-based estimate transitions from changing trends in the VI time series and are mainly used for near real-time crop monitoring \cite{gao2020a, gao2020b}. The priori curve-based approaches combine the reference curve from previous years and current observations to map LSP for the current year \cite{sakamoto2010, zhang-goldberg2012, liu-zhang2018}. In this paper, we use two approaches to evaluate LSP dynamics in drylands: a threshold and curvature model. The threshold-based approach is an established methodology, is simple to implement, and has been used to produce LSP data products from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Resolution Imaging Spectroradiometer (MODIS), and the Harmonized Landsat and Sentinel-2 (HLS) data \cite{jonsson2004, gray2019, bolton2020}. The curvature-based approach does not depend on a-prior thresholds and can capture transition dates with relatively small changes in the VI timeseries, and has been used to produce Collection 5 MODIS phenology and VIIRS phenology \cite{zhang-friedl2003, zhang-liu2018}.

For the threshold method SOS (EOS) was estimated as the DOY when the smoothed VI curve exceeded (fell below) a 10\% threshold of the relative maximum VI value. For the curvature method SOS (EOS) was estimated as the DOY when the smoothed VI curve had the highest (lowest) rate of change. The peak estimate was the day of year (DOY) of the maximum smoothed VI value, thus was independent of the SOS/EOS method. For the MAE of the in simulated study the true metrics are derived using the respective method of the original $VI_{veg}$ annual curve. Since we are using a simulated VI it can be interpreted as either EVI or NDVI. Peak was the DOY of maximum canopy greenness. All code for this analysis is available in the Zenodo data repository (https://doi.org/10.5281/zenodo.4777207).

@@

Increasing the uncertainty dampened the effect of increasing fractional vegetation cover. When the uncertainty SD was 0.04, and combined with low fractional vegetation cover, a larger proportion of simulation runs exceeded the $VI_{pixel}$ 0.1 threshold, with the reverse happening at higher fractional vegetation cover (Figure 2). Thus, VI uncertainty propagates into uncertainty in detecting dryland LSP. For example a pixel with low vegetation cover may, in some years, have seemingly detectable LSP due solely to chance instead of significantly higher productivity.

Even with a phenological pattern that is consistently detectable, the combined effects of soil background, VI uncertainty, and methodology still cause some level of bias in the estimated phenological metrics. For example, given a plant with an $VI_{veg}$ amplitude of 0.2 and 60\% fractional cover, all phenological metrics had mean average error (MAE) of at least 10 (and up to 50) days in our simulations using a 10\% of maximum threshold (Figure 3). The MAE of estimates improved with increasing vegetation cover, with increasing $VI_{veg}$ amplitude, and from using the curvature instead of threshold-based method for SOS and EOS estimation. The highest errors were seen with low $VI_{veg}$ amplitude and high (0.04) VI uncertainty. Peak greenness was more accurate than SOS or EOS in most instances when using a threshold method, especially when the $VI_{veg}$ amplitude was low (0.4 or 0.1). The curvature method outperformed the threshold method in all instances except one, when estimating EOS with a $VI_{veg}$ amplitude of 0.1. The curvature method was particularly accurate in estimating SOS of this simulated data, especially when fractional vegetation cover was low. Increasing the VI uncertainty from 0.01 to 0.04 increased the MAE of estimates in all cases regardless of fractional cover, $VI_{veg}$ amplitude, or methodology.

@@

Annual mesquite NDVI exceeded a 0.1 minimum amplitude consistently when fractional vegetation cover was between 25\% and 60\%, depending on the simulated pixel size (Figure 4). The discrepancy with pixel size is similar to what is seen with increasing VI uncertainty in the simulation analysis, but here reflects the high variation in NDVI within the mesquite canopy (Figure S2, S3). For example, given a single mesquite shrub 8m in diameter, 2m pixels randomly located within the canopy can have a wide range of NDVI values for a single acquisition date, resulting in varying amplitude given the same fractional cover. Larger pixel sizes essentially have larger samples to better estimate the mean NDVI of canopies. This leads to a scale dependence for LSP detectability at these extremely fine-scale spatial resolutions. At the 16m pixel size the NDVI variation is reduced enough to result in a relationship between fractional vegetation cover and LSP detectability which most reflects our simulated results.

As in the simulation study, errors of LSP derived transition dates decreased with increasing vegetation cover for all metrics (Figure 5). Because of the scale effect from canopy variation in NDVI, smaller pixel sizes had large variation in error rates for any given fractional cover. Using pixel sizes of 8 or 16m the MAE stabilized at approximately 20\% mesquite cover for SOS and EOS using the threshold method. The curvature method produced slightly more accurate estimates when mesquite fractional cover was less than 20\%, but equivalent or less accurate estimates at high mesquite cover. The MAE for peak estimates, which was independent to the threshold or curvature methods, was never below 40 days and commonly greater than 60. This suggests a dissimilarity between peak greenness from UAV imagery versus in-situ observations.

@@

With simulated VI curves we showed the limitations in detectability of dryland LSP with respect to fractional vegetation cover, VI amplitude, and VI uncertainty. The required fractional vegetation cover to detect phenology depends on the underlying plant traits, with evergreen plants needing higher cover than non-evergreen plants. Even when the seasonal amplitude of a pixel exceeds a minimum QA threshold, phenological metrics of sparsely vegetated drylands can still have considerable error due to other factors.  We used UAV imagery of a deciduous shrubland to validate these results, and also highlighted how within canopy VI variation affects LSP detection in high resolution imagery. The algorithm used for transition date estimation affected results as well, with the curvature method outperforming the threshold method with simulated data, and the threshold method having better or equivalent results than the curvature method with UAV imagery time series.

@@

The algorithms used for LSP detection in drylands need further study, ideally to identify the underlying mechanism which makes one method more suitable than others. Using simulated, and idealized data we found a curvature method generally outperformed a threshold method except for EOS estimates of low $VI_{veg}$ amplitudes. This was likely the result of the gradual greendown in underlying double sigmoid used in the simulation analysis, where after adding VI uncertainty the steepest part of which was highly variable. With UAV imagery the threshold method produced lower MAE than the curvature method in most instances. This could be due to the low temporal density of only eight UAV flight dates throughout the year. With limited sample size the date of maximum change in the smoothed VI curve can be more variable than the date when a relative threshold is reached. Thus, any evaluation of algorithms for dryland LSP detection must also consider the temporal scale of the sensor in addition to the vegetation and VI attributes. The most suitable algorithm for UAV imagery in a particular ecosystem will not necessarily be the most suitable for satellite based imagery in the same ecosystem.

@@

\subsection{On low amplitude errors}
Here we have highlighted a little studied aspect of LSP research. As $VI_{pixel}$ amplitude decreases, either from lower fractional vegetation cover or lower $VI_{veg}$ amplitude, potential error in transition date estimates increase. This is associated with the inherent parameter uncertainty in smoothing algorithms, leading to a statistical limitation in estimating transition dates. We illustrated this effect in Figure 8, where three VI curves have the same VI uncertainty (SD = 0.01) but different $VI_{pixel}$ amplitudes. As amplitude decreases, the potential range of onset dates increases. This happens with both a threshold- and curvature-based approach, though the curvature-based approach is less sensitive. As seen in both the simulation and UAV imagery analysis here this leads to higher error potential at lower $VI_{pixel}$ amplitudes. Confidence intervals, and errors of resulting transition date estimates, can be decreased with different temporal composite methods and/or with smaller temporal resolutions \cite{zhang-friedl2009}. A best practice would also be to incorporate transition date confidence intervals into LSP studies for more appropriate statistical tests.

@@

In areas where $VI_{veg}$ amplitude is extremely low the timing of peak greenness is likely the most suitable metric. Here we found estimates of peak greenness showed less bias than those for the start or end of season, especially when $VI_{veg}$ amplitude was low. Other studies have shown the timing and amplitude of peak greenness to be reliable in studying evergreen phenology \cite{walker2014, walker2015} and annual vegetation biomass \cite{casady2013}. Though caution must still be observed here since low $VI_{veg}$ amplitude can lead to improbable winter peaks \cite{norris2020}.

@@

Similarly, UAV imagery can also be used to track phenology at the individual plant level. We found a potential shortcoming in using UAV imagery for LSP analysis, where the peak estimate had large errors relative to SOS and EOS estimates. This could be due to the true date of peak greenness being between our flight dates. Another possible cause is variation in timing of leaf maturity and senescence within the plant canopy, causing differences due to the oblique angle or categorical ocular estimates of in-situ observations compared to UAV imagery. Understory species within the mesquite canopy may drive an earlier peak when measured by UAV imagery. Thus when using UAV imagery to estimate phenology, one should maximize the number of sample dates, especially since peak greenness can only be estimated retrospectively. As high resolution imagery becomes more accessible and commonly used, the relationship between fine scale canopy attributes and how they relate to in-situ observations should be explored more \cite{klosterman2018, cheng-vrieling2020}.

@@

Drylands constitute approximately 41\% of the Earth’s surface \cite{hoover2020} and their high variability in annual productivity plays an important role in global terrestrial net primary productivity \cite{ahlstrom2015}. For these reasons, it is important to understand underlying factors limiting LSP studies in drylands. We found that the interaction of fractional vegetation cover and seasonal VI amplitude of plant canopies are the primary drivers of dryland LSP detectability and accuracy. At intermediate levels of fractional vegetation cover VI uncertainty associated with shadows, atmospheric interference, and view angle affects detectability. Retrospective studies will always be limited in their ability to measure LSP in drylands due to coarse resolution sensors. Full uncertainty propagation, from sensor measurement to statistical models, can help contextualize results in any LSP study, but is especially pertinent for drylands. High resolution sensors are the most promising path forward for measuring LSP in drylands since sparse vegetation is advantageous for identification of individual plants. Drylands thus have the potential to become an exemplary environment for future LSP research.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{setspace}
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{lineno}
\linenumbers

@@

%%%%%% Abstract %%%%%%
\begin{abstract}
Land surface phenology (LSP) enables global scale tracking of ecosystem processes, but its utility is limited in drylands due to low vegetation cover and resulting low amplitudes of vegetation indices (VIs). Drylands play an important role in global biodiversity, food security, and the carbon cycle. Thus it is important to understand the limitations in detecting and measuring dryland dynamics.

Here, using simulated data and multi-temporal UAV imagery of a desert shrubland, we explore the feasibility of detecting LSP with respect to fractional vegetation cover, plant functional types, VI uncertainty, and different detection algorithms. We found that plants with distinct VI signals, such as deciduous shrubs, require at least 30-40\% fractional cover to consistently detect LSP. Evergreen plants, with lower seasonal VI amplitude, require considerably higher cover and can have undetectable phenology even with 100\% vegetation cover. Our evaluation of two algorithms showed that neither performed the best in all cases. Even with adequate cover, biases in phenological metrics can still exceed 20 days, and can never be 100\% accurate due to VI uncertainty from shadows, sensor view angle, and atmospheric interference.

We showed how high-resolution UAV imagery enables LSP studies in drylands, and highlighted important scale effects driven by within canopy VI variation. With high-resolution imagery the open canopies of drylands are beneficial as they allow for straightforward identification of individual plants, enabling the tracking of phenology at the individual level. Drylands thus have the potential to become an exemplary environment for future LSP research.

@@

Land surface phenology (LSP) enables ecosystem scale tracking of the drivers and consequences of a changing climate. Satellite sensor-derived vegetation indices (VIs) track the progression of green vegetation throughout the year, and from this time series the seasonal transitions between the dormant and growing seasons are derived \cite{zhang-liu2018, friedl2006, jonsson2004, sakamoto2010, gao2017, gray2019, bolton2020, zeng-wardlow2020}. Numerous studies measure the short- and long-term LSP trends, linking them to drivers such as weather and climate, land cover change, and disturbances \cite{piao-wang2020}. Yet the use of LSP is limited outside regions with adequate vegetation cover and plants with distinct seasonal change. Temperate deciduous forests have the most distinct LSP signal, while in other vegetation types LSP is difficult to discern due to combinations of low signal-to-noise ratio between growing and dormant season VI, high snow cover, and low vegetation cover \cite{richardson2018b, white2009}.

In dryland ecosystems low vegetation cover is the primary limitation in detecting land surface phenology \cite{ma2013}. High soil fractional cover can cause the growing season VI to be indistinguishable from the dormant season VI, making phenology extraction impossible. Additionally, some areas with adequate vegetation cover are dominated by evergreen vegetation, thus can have a growing season VI too low to reliably detect seasonal transitions \cite{walker2015}. Even in areas with high vegetation cover and plants with distinct signals, phenology may be impossible to detect some years due little to no plant productivity from inadequate precipitation. These limitations cause drylands to be excluded in many large scale LSP analyses \cite{jeong2011, zhu2012, shen-zhang2014, xin-broich2015, garonna2016}, while other analysis include them without any regard for their detectability and resulting bias \cite{buitenwerf2015, VanLeeuwen2010}. Most studies focusing on drylands evaluate aggregate or peak annual VI as opposed to distinct seasonal transitions \cite{walker2015, hou-du2019}, and even then can occasionally have inconclusive results \cite{fensholt2012, andela2013}.

Drylands cover 41\% of the Earth’s surface \cite{hoover2020} and reliable measurement of LSP in drylands are needed in understanding these ecosystems and the large-scale roles they play in global processes. Drylands account for much of the long term trends in increased carbon uptake globally \cite{ahlstrom2015} and leaf phenology drives much of the inter-annual variation in water and carbon uptake in dryland systems \cite{whitley2017, renwick2019}. One quarter of terrestrial vertebrates utilize drylands to some degree \cite{safriel2005} and changing plant phenology can have cascading effects on primary consumers and higher trophic interactions \cite{aiken2017}. Additionally, dryland rangelands support 35-84\% of livestock globally, thus are vital source of ecosystem services \cite{chapin1996}.

Here we explore key attributes of VI time series and how they affect reliable detection of LSP in dryland ecosystems. We use a spectral mixing model to simulate LSP with varying ranges of plant fractional cover, seasonal VI amplitude, and VI uncertainty to determine the feasibility of detecting LSP in different scenarios and with different detection algorithms. We then use high spatial resolution, multi-temporal unmanned aerial vehicle (UAV) imagery to verify some of these relationships. Finally we discuss the feasibility of detecting dryland LSP currently and in the future.

@@

Here we explore how variation in the three components in Eq. 1 (i.e., ranges of plant fractional cover, seasonal VI amplitude, and VI uncertainty) affect the ability to detect a distinct phenological signal in sparsely vegetated areas. We start with an idealized annual VI curve derived from a double sigmoid model to represent the VI of single plant canopy (Figure 1, \cite{elmore2012}). The curve’s amplitude (i.e., the difference between dormant and peak VI values) reflects a combination of plant functional type and leaf area index (LAI). The seasonal amplitude of non-evergreen plants is driven by LAI up to an LAI of approximately 2, while evergreen plants have little to no seasonal amplitude \cite{gamon1995, carlson1997}. Higher LAI will generally lead to higher amplitudes, with this effect being more prominent in deciduous plants than evergreen plants \cite{baghzouz2010, smith2019}. The leaves of senesced grasses have distinctly lower VI than green leaves, thus are expected to behave similarly to deciduous plants \cite{huete1987}.

@@

For the simulation analysis we repeated this process while varying the three different factors from Eq. 1: 1) the fractional vegetation cover ($Cover_{veg}$), using values from 0-100\%, 2) the amplitude of $VI_{veg}$ from dormant to peak, adjusted using a parameter in the double sigmoid model, and 3) the size of the error term $\varepsilon$ using values of 0.01 and 0.04 for the standard deviation of the gaussian noise. For each combination we repeated the process 100 times. A low growing season VI relative to the dormant season is one of the main challenges in dryland LSP \cite{smith2019}. Phenological extraction methods account for this in several ways. A common approach is discarding pixels which do not exceed a threshold of 0.1 VI units between the dormant season and peak growth (i.e., a minimum $VI_{pixel}$ amplitude of 0.1; \cite{white2009, jeong2011, garonna2016, gray2019}), while others have used statistical power methods of the time-series to filter out pixels with low signal-to-noise \cite{dannenberg2017}. Here we evaluate the detectability of LSP in two ways: 1) The frequency at which the $VI_{pixel}$ meets or exceeds an amplitude of 0.1 VI units, and 2) mean absolute error (MAE) in the estimated phenological metrics start of season (SOS), end of season (SOS), and peak of season (POS).

\subsection{UAV Study Area, Acquisition, and Processing}

@@

@@

Using this time series of NDVI rasters we developed a sampling routine to test the detectability of phenology with respect to fractional vegetation cover similar to the simulation study. We randomly placed 12,000 simulated pixels across the site with sizes of 2, 4, 8, or 16 meters. We then extracted the average NDVI within each one for each image date. This random placement allowed us to produce 12,000 annual NDVI time series representing a range of fractional cover for honey mesquite across several pixel resolutions. The average shrub diameter is approximately 8m, so multiple pixel sizes allow us to test both L- and H-resolution LSP detection \cite{strahler1986}. As in the simulation study, each annual time series was fit with a cubic smoothing spline and we then calculated whether the smoothed time series exceeded a 0.1 NDVI threshold and extracted phenology metrics using a 10\% of relative max threshold method and the change rate method as described below.

@@

\subsection{Detection Methods and Analysis}
Many remote sensing algorithms have been developed in recent decades for estimating transition dates from VI timeseries, and can be categorized in four broad categories: threshold, curve-based, trend, and priori curve-based approach \cite{gao-zhang2021, zeng-wardlow2020}. Here we use two approaches to evaluate LSP dynamics in drylands: a threshold and curve-based model. The threshold approach uses either fixed or dynamic thresholds of VI values to identify transition dates, is an established and simple methodology, and has been used to produce phenology data products from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Resolution Imaging Spectroradiometer (MODIS), and the Harmonized Landsat and Sentinel-2 (HLS) data \cite{jonsson2004, gray2019, bolton2020}. A curve-based approach detects the inflection points that show the quickest changes in the VI time series, thus does not depend on a priori thresholds. Curve-based approaches can capture transition dates with relatively small changes in the VI timeseries, and have been used to produce Collection 5 MODIS phenology and VIIRS phenology \cite{zhang-friedl2003, zhang-liu2018}. Both of these approaches are commonly used in LSP studies \cite{garonna2016, jeong2011, buitenwerf2015}.

For the threshold method SOS (EOS) was estimated as the DOY when the smoothed VI curve exceeded (fell below) a 10\% threshold of the relative maximum VI value. For the curve-based method SOS (EOS) was estimated as the DOY when the smoothed VI curve had the highest (lowest) rate of change using the numerically calculated first derivative \cite{tateishi2004}. The POS estimate was the DOY of the maximum smoothed VI value, thus was independent of the SOS/EOS method. For the MAE of the simulation study the true metrics are derived using the respective method of the original $VI_{veg}$ annual curve. Since we are using a simulated VI it can be interpreted as either EVI or NDVI. All code for this analysis is available in the Zenodo data repository (https://doi.org/10.5281/zenodo.4777207).

@@

Increasing the uncertainty dampened the effect of increasing fractional vegetation cover. When the uncertainty SD was 0.04, and combined with low fractional vegetation cover, a larger proportion of simulation runs exceeded the $VI_{pixel}$ 0.1 threshold, with the reverse happening at higher fractional vegetation cover (Figure 2). At low and high fractional vegetation cover VI uncertainty had little to no effect on whether the 0.1 threshold was exceeded. Thus, VI uncertainty propagates into uncertainty in detecting dryland LSP, with this effect being highest at intermediate levels of fractional vegetation cover. For example a pixel with low vegetation cover may, in some years, have seemingly detectable LSP due solely to chance instead of significantly higher productivity.

Even with a phenological pattern that is consistently detectable, the combined effects of soil background, VI uncertainty, and methodology still cause some level of bias in the estimated phenological metrics. For example, given a plant with an $VI_{veg}$ amplitude of 0.2 and 60\% fractional cover, all phenological metrics had mean average error (MAE) of at least 10 (and up to 50) days in our simulations using a 10\% of maximum threshold (Figure 3). The MAE of estimates improved with increasing vegetation cover, with increasing $VI_{veg}$ amplitude, and from using the change rate instead of threshold-based method for SOS and EOS estimation. The highest errors were seen with low $VI_{veg}$ amplitude and high (0.04) VI uncertainty. POS estimates were more accurate than SOS or EOS in most instances when using a threshold method, especially when the $VI_{veg}$ amplitude was low (0.4 or 0.1). The change rate method outperformed the threshold method in all instances except one, when estimating EOS with a $VI_{veg}$ amplitude of 0.1. The change rate method was particularly accurate in estimating SOS of this simulated data, especially when fractional vegetation cover was low. Increasing the VI uncertainty from 0.01 to 0.04 increased the MAE of estimates in all cases regardless of fractional cover, $VI_{veg}$ amplitude, or methodology.

@@

Annual mesquite NDVI exceeded a 0.1 minimum amplitude consistently when fractional vegetation cover was between 25\% and 60\%, depending on the simulated pixel size (Figure 4). Larger pixel sizes require less fractional cover to consitently exceed the 0.1 amplitude. this discrepancy with pixel size is similar to what is seen with increasing VI uncertainty in the simulation analysis, but here reflects the high variation in NDVI within the mesquite canopy. For example, given a single mesquite shrub 8m in diameter, 2m pixels randomly located within the canopy can have a wide range of NDVI values for a single acquisition date, resulting in varying amplitude given the same fractional cover (Figure S2, S3). Larger pixel sizes essentially have larger samples to better estimate the mean NDVI of canopies. This leads to a scale dependence for LSP detectability at these extremely fine-scale spatial resolutions. At the 16m pixel size the NDVI variation is reduced enough to result in a relationship between fractional vegetation cover and LSP detectability which most reflects our simulated results.

As in the simulation study, errors of LSP derived transition dates decreased with increasing vegetation cover for all metrics (Figure 5). Because of the scale effect from canopy variation in NDVI, smaller pixel sizes had large variation in error rates for any given fractional cover. Using pixel sizes of 8 or 16m the MAE stabilized at approximately 20\% mesquite cover for SOS and EOS using the threshold method. The change rate method produced slightly more accurate estimates when mesquite fractional cover was less than 20\%, but equivalent or less accurate estimates at high mesquite cover. The MAE for POS, which was independent to the threshold or change rate methods, was never below 40 days and commonly greater than 60 regardless of pixel size. This suggests a dissimilarity between POS from UAV imagery versus in-situ observations.

@@

With simulated VI curves we showed the limitations in detectability of dryland LSP with respect to fractional vegetation cover, VI amplitude, and VI uncertainty. The required fractional vegetation cover to detect phenology depends on the underlying plant traits, with evergreen plants needing higher cover than non-evergreen plants. For example our UAV analysis showed that 20-30\% deciduous shrub cover is needed to consistently detect LSP, while Peng et al. 2021 \cite{peng2021} found that approximately 50\% cover is needed for consistent detection in evergreen dominated shrublands. Even when the seasonal amplitude of a pixel exceeds a minimum QA threshold, phenological metrics of sparsely vegetated drylands can still have considerable error due to other factors.  Our UAV imagery analysis validated these results, and also highlighted how within canopy VI variation affects LSP detection in high-resolution imagery. The algorithm used for transition date estimation affected results as well, with the change rate method outperforming the threshold method with simulated data, and the threshold method having better or equivalent results than the change rate method with UAV imagery time series.

@@

The algorithms used for LSP detection in drylands need further study, ideally to identify the underlying mechanism which makes one method more suitable than others. Using simulated, and idealized data we found the change rate method generally outperformed the threshold method except for EOS estimates of low $VI_{veg}$ amplitudes. This was likely the result of the gradual greendown in underlying double sigmoid used to generate the simulated data, where after adding VI uncertainty the steepest part of which was highly variable in the resulting cubic spline smoother. With UAV imagery the threshold method produced lower MAE than the change rate method in most instances. This could be due to the low temporal density of only eight UAV flight dates throughout the year. With limited sample size the date of maximum change in the smoothed VI curve can be more variable than the date when a relative threshold is reached. Thus, any evaluation of algorithms for dryland LSP detection must also consider the temporal scale of the sensor in addition to the vegetation and VI attributes. The most suitable algorithm for UAV imagery in a particular ecosystem will not necessarily be the most suitable for satellite based imagery in the same ecosystem.

@@

\subsection{On low amplitude errors}
Here we have highlighted a little studied aspect of LSP research. As $VI_{pixel}$ amplitude decreases, either from lower fractional vegetation cover or lower $VI_{veg}$ amplitude, potential error in transition date estimates increase. This is associated with the inherent parameter uncertainty in smoothing algorithms, leading to a statistical limitation in estimating transition dates. We illustrated this effect in Figure 8, where three VI curves have the same VI uncertainty (SD = 0.01) but different $VI_{pixel}$ amplitudes. As amplitude decreases, the potential range of onset dates increases. This happens with both a threshold and change rate method, though the change rate method is less sensitive. As seen in both the simulation and UAV imagery analysis here this leads to higher error potential at lower $VI_{pixel}$ amplitudes. Confidence intervals, and errors of resulting transition date estimates, can be decreased with different temporal composite methods and/or with smaller temporal resolutions \cite{zhang-friedl2009}. A best practice would also be to incorporate transition date confidence intervals into LSP studies for more appropriate statistical tests.

@@

In areas where $VI_{veg}$ amplitude is extremely low the timing of POS is likely the most suitable metric. Here we found estimates of POS showed less bias than those for the SOS or EOS, especially when $VI_{veg}$ amplitude was low. Other studies have shown the timing and amplitude of POS to be reliable in studying evergreen phenology \cite{walker2014, walker2015} and annual vegetation biomass \cite{casady2013}. Though caution must still be observed here since low $VI_{veg}$ amplitude can lead to improbable winter peaks \cite{norris2020}.

@@

Similarly, UAV imagery can also be used to track phenology at the individual plant level. We found a potential shortcoming in using UAV imagery for LSP analysis, where the POS had large errors relative to SOS and EOS estimates. This could be due to the true date of POS being between our flight dates. Another possible cause is variation in timing of leaf maturity and senescence within the plant canopy, causing differences due to the oblique angle or categorical ocular estimates of in-situ observations compared to UAV imagery. Understory species within the mesquite canopy may drive an earlier peak when measured by UAV imagery. Thus when using UAV imagery to estimate phenology, one should maximize the number of sample dates, especially since POS can only be estimated retrospectively. As high resolution imagery becomes more accessible and commonly used, the relationship between fine scale canopy attributes and how they relate to in-situ observations should be explored more \cite{klosterman2018, cheng-vrieling2020}.

@@

Drylands constitute approximately 41\% of the Earth’s surface \cite{hoover2020} and their high variability in annual productivity plays an important role in global terrestrial net primary productivity \cite{ahlstrom2015}. Drylands also account for large amounts of ecosystem services and biodiversity \cite{safriel2005, chapin1996}. Many of these aspects are driven by or correlated with LSP, thus it is important to understand the drivers and limitations of reliable LSP detection in drylands.

As in other studies we found that the interaction of fractional vegetation cover and seasonal VI amplitude of plant canopies are the primary drivers of dryland LSP detectability and accuracy \cite{smith2019, peng2021}. High vegetation cover leads to higher detectability, except when plants have low VI amplitudes, for example in evergreen vegetation. VI uncertainty associated with shadows, atmospheric interference, and view angle is largely negligible in determining if a VI signal will exceed a minimum amplitude when fractional vegetation cover is either very high or very low. At intermediate levels of fractional vegetation cover VI uncertainty can increase the false positives and negatives of the true VI signal exceeding a minimum amplitude. Regardless of whether a VI signal exceeds a minimum threshold, reducing VI uncertainty can reduce the errors of resulting LSP metrics in all cases. Of the two algorithms used here for LSP estimates, neither performed the best in all scenarios. Studies evaluating the numerous LSP methodologies with a focus on drylands would be highly beneficial, especially if they could identify underlying factors which determine which method performs best in a particular plant community.

High resolution sensors with sub-meter resolution are a promising path forward for measuring LSP in drylands. These sensors overcome the problem imposed by low vegetation cover, allowing for LSP detection of individual plants. Frequent visits are still needed to adequately capture all seasonal dynamics, which may be problematic when using labor intensive UAVs. High resolution satellite based sensors will be highly suitable for LSP studies in drylands due to frequent visits and low cloud cover in dryland environments. Since high resolution imagery also allows for the identification of individual plants, it will be possible to measure LSP at the individual level. Drylands thus have the potential to become an exemplary environment for future LSP research.
